---
title: "Minimum Viable Product: Princeton Gerrymandering Project"
author: "Ochuwa Garuba"
date: "`r format(Sys.time(), 'Compiled %X on %A, %B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
header-includes:
- \usepackage{fvextra}
- \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---
# Data Set Up

```{r set up, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Load packages
library(ggplot2)
library(tidyverse)
library(GGally)

# Set the seed to create reproducible code
set.seed(123)

# Load data
elections <- read.csv('/Users/ochuwagaruba/Desktop/Vanderbilt University/Spring 2026/HIST 1354/Data/congressional_election_results_post1948.csv')
```

# Data Wrangling

```{r data wrangling, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Clean data
southern_states <- c(
  "AL", "GA", "LA", "MS",
  "SC", "TX", "VA",
  "NC", "FL", "TN"
)

# Filter the data
elections_filtered <- elections[elections$State %in% southern_states, ]
```

# Data Inspection

```{r data inspection, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Check data
str(elections_filtered)
head(elections_filtered)
summary(elections_filtered)
```

# Predictive Modeling

```{r modeling, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
# Prep categorical data for the linear regression model
model_data <- elections_filtered %>%
  mutate(
    State = as.factor(State),
    Incumbent_Party = as.factor(ifelse(Incumbent == 1, "Democrat", "Republican"))
  )

# Run the linear regression
election_model <- lm(D.Voteshare ~ Year + State + Incumbent_Party, data = model_data)

# View results
summary(election_model)
```

# Line Graph 1: Historical Trends in the South

```{r line graph 1, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
ggplot(elections_filtered, aes(x = Year, y = D.Voteshare, color = State)) +
  geom_smooth(se = FALSE, size = 0.8) +
  geom_vline(xintercept = c(1965), linetype = "dotted", alpha = 0.5) +
  facet_wrap(~State) +
  labs(
    title = "Southern Realignment: State-by-State Comparison",
    subtitle = "Dotted lines indicates 1965, which is when the Voting Rights Act passed"
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(legend.position = "none")
```

# Table: Calculate average vote share by state and decade

```{r table, echo = TRUE, eval = TRUE, comment = NA, warning = FALSE, message = FALSE}
decade_summary <- elections_filtered %>%
  mutate(Decade = floor(Year / 10) * 10) %>%
  group_by(State, Decade) %>%
  summarise(Avg_Voteshare = mean(D.Voteshare, na.rm = TRUE), .groups = 'drop') %>%
  pivot_wider(names_from = Decade, values_from = Avg_Voteshare)
decade_table_clean <- decade_summary %>%
  mutate(across(where(is.numeric), ~ paste0(round(.x * 100, 1), "%")))
knitr::kable(decade_table_clean, caption = "Average Democratic Vote Share by State and Year")
```

# Reflection

The biggest technical hurdle was narrowing down which variables to include in my linear regression model. While my dataset from the Princeton Gerrymandering Project contains several columns, I had to be intentional about which factors would actually predict a relevant historical outcome. I initially considered including raw vote totals or the specific district numbers, but I  ralized that these would complicate the model without adding real value to the overarching narrative. Districts are redrawn every ten years, making District 1 a moving target geographically. Vote totals are also the raw ingredients of the vote share itself. Deciding to focus on Year, State, and Incumbency required a deliberate shift from simply using all available data to selecting variables that represent the actual engines of Southern political change: time, location, and political power. Based on this preliminary engine run, the data strongly supports my initial hypothesis regarding the realignment of the American South. The model shows a clear, statistically significant decline in Democratic vote share over time, with an average annual drop of about 28 percent.

The most striking finding was the massive influence of incumbency. The model indicates that a Republican incumbent reduces the expected Democratic vote share by roughly 47 percent, which highlights how the party in power can anchor a district's political outcome Furthermore, the state-by-state breakdown reveals that while the entire South moved away from the Democratic Party, the starting points and slopes differed. For instance, Tennessee showed much more early-stage competition in the 1940s and 50s than states in the Deep South like Mississippi. This suggests that there are nuanced regional tensions between federal oversight and local political shifts.